Mapping mutational effects along the evolutionary landscape of HIV envelope

### 2017-11-30 paper/manuscript.tex
finalized tree figure, legend, supplements

But interpreting these studies in the context of Env evolution requires addressing a fundamental question: How informative are mutational studies of a single protein variant about constraints on long-term evolution?
During protein evolution, substitutions at one site can shift the effect of mutations at other sites~\citep{natarajan2013epistasis,gong2013stability,harms2014historical,podgornaia2015pervasive,starr2016epistasis}.
Such shifts can accumulate as substitutions become entrenched via epistatic interactions with subsequent changes~\citep{starr2017pervasive,pollock2012amino,shah2015contingency} -- although the magnitude of these shifts is usually limited~\citep{doud2015site,chan2017correlation,ashenberg2013mutational,risso2014mutational}.
Given that the Envs of circulating HIV strains represent a vast collection of homologs that often differ at $>$100 residues, such shifts could make the outcome of any study highly dependent on the Env used.
The extent to which this is actually the case is unclear, since the few protein-wide studies of mutational shifts have examined proteins far simpler than Env, which forms a large heterotrimeric complex that transitions through multiple conformational states~\citep{munro2014conformational,ozorowski2017open}.

Here we measure the effects on viral replication of all single amino-acid mutations to the ectodomains of two transmitted-founder virus Envs that differ by $>$100 mutations.
By comparing these complete mutational maps, we identify sites where amino-acid preferences have shifted between the Envs.
Most sites show no detectable shifts, but 30 sites have clearly altered their amino-acid preferences.
These shifts tend to involve sites that strongly prefer a specific amino acid in one Env but tolerate many amino acids in the other.
The shifted sites cluster in Env's structure, but are often distant from any of the amino-acid substitutions that distinguish the two Envs -- demonstrating the action of long-range epistasis.
By aggregating our measurements for both Envs, we identify sites that evolve faster or slower in nature than expected given the functional constraints measured in the lab.
Overall, our work provides across-strain genotype-to-phenotype maps that complement existing work characterizing Env's diversity at the level of protein sequence and structure.

@@

\subsection*{Deep mutational scanning of two Env homologs from transmitted-founder viruses}
Correlations between replicates in prior studies:
\citet{haddox2016experimental} all sites, had $R = 0.45$ to $R = 0.50$;

@@

\begin{figure}
\centerline{\bf \Large A \hspace{0.52\textwidth} \bf B \hspace{0.35\textwidth}}
\centerline{\includegraphics[width=0.64\textwidth]{figures/tree_plot.pdf} \includegraphics[width=0.36\textwidth]{figures/masked_alignment_identity.pdf}}
\caption{\label{fig:tree}
CAPTION
}
\figdata{The coding sequence alignment of clade A \textit{env} sequences in FASTA format is in \texttt{cladeA\_alignment.fasta}.}
\figdata{Sites masked in all phylogenetic analyses because they were not mutagenized in our experiments or are poorly alignable are listed in \texttt{alignment\_mask.csv}.}
\end{figure}

>>>>

But interpreting these studies in the context of Env evolution requires addressing a fundamental question: How informative are mutational studies of a single protein variant about constraints on long-term evolution?
During protein evolution, substitutions at one site can shift the effect of mutations at other sites~\citep{natarajan2013epistasis,gong2013stability,harms2014historical,podgornaia2015pervasive,starr2016epistasis}.
Such shifts can accumulate as substitutions become entrenched via epistatic interactions with subsequent changes~\citep{starr2017pervasive,pollock2012amino,shah2015contingency} -- although the magnitude of these shifts is usually limited~\citep{doud2015site,chan2017correlation,ashenberg2013mutational,risso2014mutational}.
Given that the Envs of circulating HIV strains represent a vast collection of homologs that often differ at $>$100 residues, such shifts could make the outcome of any study highly dependent on the Env used.
The extent to which this is actually the case is unclear, since the few protein-wide studies of mutational shifts have examined proteins that are structurally far simpler than Env, which forms a large heavily glycosylated heterotrimeric complex that transitions through multiple conformational states~\citep{munro2014conformational,ozorowski2017open}.

Here we measure the effects on viral replication of all single amino-acid mutations to the ectodomains of two transmitted-founder virus Envs that differ by $>$100 mutations.
By comparing these complete mutational maps, we identify sites where amino-acid preferences have shifted between the Envs.
Most sites show no detectable shifts, but 30 sites have clearly altered their amino-acid preferences.
These shifted sites usually strongly prefer a specific amino acid in one Env but have shifted to tolerate many amino acids in the other Env.
The shifted sites cluster in Env's structure, but are often distant from any amino-acid substitutions that distinguish the two Envs -- demonstrating the action of long-range epistasis.
By aggregating our measurements for both Envs, we identify sites that evolve faster or slower in nature than expected given the functional constraints measured in the lab.
Overall, our work provides across-strain genotype-to-phenotype maps that inform the analysis of Env's structure and evolution.

@@

\subsection*{Two Env homologs from clade A transmitted-founder viruses}

\begin{figure}
\centerline{\includegraphics[clip=true, trim= 0in 0in 1in 0in, angle=-90, width=0.9\textwidth]{figures/tree_plot.pdf}}
\caption{\label{fig:tree}
Phylogenetic tree showing the relationship of BG505 and BF520 to other clade A Envs.
The tree shows the 69 Envs in the alignment in Figure~\ref{fig:tree}-Figure~data~\ref{figdata:alignment}, which is a subsample of substantially unique sequences clade A sequences from the group M alignment in the Los Alamos HIV sequence database (\url{http://www.hiv.lanl.gov}).
Sites not mutagenized in our experiments (the signal peptide and cytoplasmic tail) or that are poorly alignable (mostly sites in the variable loops) were masked as indicated in Figure~\ref{fig:tree}-Figure~data~\ref{figdata:alignment}, leaving 616 alignable sites.
The pairwise identity of BG505 and BF520 to the other sequences at these alignable sites is shown in Figure~\ref{fig:tree}-Figure~data~\ref{figsupp:identity}.
The tree topology was inferred using \texttt{RAxML}~\citep{stamatakis2014raxml} under the GTRCAT model of nucleotide substitution, and then branch lengths were optimized under the M0 version of the Goldman-Yang model~\citep{yang2000codon} using \texttt{phydms}~\citep{hilton2017phydms}.
}
\figsupp[\label{figsupp:identity}  Pairwise identity of all Env sequences in our clade A alignment to BG505 and BF520 after masking poorly alignable or non-mutagenized sites.]
{The histograms show the pairwise amino-acid identity of each Env to all other sequences in the clade A alignment in Figure~\ref{fig:tree}-Figure~data~\ref{figdata:alignment} after masking the sites delineated in Figure~\ref{fig:tree}-Figure~data~\ref{figdata:mask}.
There are 616 non-masked sites.
The pairwise protein identity between BG505 and BF520 is 86.2\% (721 of 836 sites identical) when considering \emph{all} sites, and 89.1\% (549 of 616 sites identical) when considering just the non-masked sites.}
{\includegraphics[width=0.7\textwidth]{figures/masked_alignment_identity.pdf}}
\figdata{\label{figdata:alignment} The sequence alignment of clade A \textit{env} coding sequences in FASTA format is in \texttt{cladeA\_alignment.fasta}.}
\figdata{\label{figdata:mask} The 240 Env sites masked in all phylogenetic analyses because they were not mutagenized in our experiments or are poorly alignable are listed in \texttt{alignment\_mask.csv}.}
\end{figure}

\subsection*{Deep mutational scanning of two Env homologs from transmitted-founder viruses}

Correlations between replicates in prior studies:
\citet{haddox2016experimental} all sites, had $R = 0.45$ to $R = 0.50$;

@@
